Simulating cosmic variance in Weak Lensing: effects on parameter inferences

### 2015-11-22 Paper/paper.tex
finalized second draft

\begin{abstract}
Constraining cosmology using Weak Gravitational Lensing consists in comparing measured features with simulated ones. An accurate estimate of the feature covariance matrix is essential to obtain accurate parameter confidence intervals. When this covariance matrix $\bb{C}$ is measured from simulations, an important question to ask is how big the simulation set should be for an accurate estimation of $\bb{C}$. We construct different mock ensembles with $N_r$ realizations of the same shear field, based on $N_s<N_r$ independent $N$--body simulations. Using the shear--shear power spectrum as a summary statistic, we find that a single $N$--body simulation can be recycled a large number of times to produce an ensemble of $N_r=10^5$ shear realizations that are mutually independent. We also find that, for a range of summary statistics that include shear--shear power spectra measured at different scales and convergence peak counts, parameter constraints degrade faster than the $O(N_b)$ Gaussian prediction by \citep{DodelsonSchneider13}, where $N_b$ is the length of the feature vector.       
\end{abstract}

@@

\subsection{Shear field simulations}
\label{shearsim}
%
In this paragraph we describe how we constructed our shear field ensembles. Background galaxies at redshift $z_s$ are lensed by large scale structure between $z=0$ and $z_s$. The shape distortions due to the cosmic shear $\pmb{\gamma}$ can be computed in terms of the dark matter gravitational potential $\Phi(\bb{x},z)$. Because the evolution of $\Phi$ in redshift is non--linear, its evolution needs to be computed with numerical simulations. We make use of the public code \texttt{Gadget2} \citep{Gadget2}, with which we run a sequence of dark--matter--only $N$--body simulations that track the evolution of the density fluctuations in the large scale structure of the universe. We assume a standard $\Lambda$CDM framework with $(\Omega_m,\Omega_\Lambda,h,w,\sigma_8,n_s)=(0.26,0.74,0.72,-1,0.8,0.96)$ and we fix the comoving size of the simulation box to $240\mathrm{Mpc}/h$. We fill the box with $512^3$ particles, which correspond to a mass resolution of $10^{10}M_\odot$. We assume a uniform galaxy distribution at a constant redshift $z_s=2$ (at which the simulation box has an angular size of $\theta_{box}=3.5\mathrm{deg}$) and we disctretize the mass distribution between $z_s$ and the observer at $z=0$ with a sequence of 46 two dimensional lenses of thickness $80\mathrm{Mpc}/h$. The surface density on each lens is computed by projecting the three dimensional density measured from \texttt{Gadget2} snapshots. We then apply the multi--lens--plane algorithm (see \citep{RayTracingHartlap,RayTracingJain} for example) to trace the deflections of $2048^2$ light rays arranged on a square grid of total size $\theta_{box}$, from $z=0$ to $z_s$. The light deflection calculations have been performed with our implementation of multi--lens--plane algorithm, which is part of the \texttt{LensTools} computing package \citep{LensTools} (released under the MIT license). Different realizations $r$ of the same shear field $\pmb{\gamma}_r(\pmb{\theta})$ can be generated with different choices of the lenses that lay between the observer and $z_s$. The randomization procedure we adopt is the following:

@@

\begin{equation}
\label{covest}
\bbh{C} = \frac{1}{N_r-1}\sum_{r=1}^{N_r} (\bbh{d}_r - \bar{\bb{d}}) (\bbh{d}_r - \bar{\bb{d}})^T
\end{equation}
%
Assuming a normal feature likelihood, together with a flat prior on the parameter space, the parameter posterior distribution $\mathcal{L}(\bb{p}\vert\bbh{d})$ can be obtained using Bayes theorem
\begin{equation}
\label{posteriorbayes}
\mathcal{L}(\bb{p}\vert\bbh{d}) = \mathcal{N}_\mathcal{L}\exp{\left[-\frac{1}{2}(\bbh{d}-\bb{d}(\bb{p}))^T\bb{C}^{-1}(\bbh{d}-\bb{d}(\bb{p}))\right]}
\end{equation}
%
For the sake of simplicity, we approximate the posterior as a Gaussian around its maximum. This is easily done approximating the simulated feature at first order around a point $\bb{p}_0$ (that ideally is the maximum of (\ref{posteriorbayes}))
\begin{equation}
\bb{d}(\bb{p}) \approx \bb{d}_0 + \bb{d}_0^\prime(\bb{p}-\bb{p}_0) 
\end{equation}
%
To measure the feature derivatives $\bb{d}'_0$ with respect to cosmology, we make use of the public code code NICAEA (see \citep{Nicaea}) for the power spectrum, and we use an independent simulation set (that contains mocks at different combinations of $(\Omega_m,w,\sigma_8)$) for the peak counts.
We can build the estimator for the posterior location parameter $\bbh{p}$, given the observation $\bbh{d}$, as follows:
%
\begin{equation}
\label{estimatormean}
\bbh{p} = \bb{p}_0 + \bbh{T}(\bbh{d}-\bb{d}_0)
\end{equation}

\begin{equation}
\label{dodelsonscaling}
\langle\h{\Sigma}_\bb{p}\rangle=\Sigma_\bb{p}(1+B(N_b-N_p))
\end{equation}
%
This is the result obtained by \citep{DodelsonSchneider13}.
\item If we are not interested in measuring the location parameter $\bbh{p}$, but just the width of the parameter posterior, we can tempted to replace the true covariance $\bb{C}$ with its estimator $\bbh{C}$. If we do that, and we do not correct for the bias (\ref{psibias}), the parameter variance will have a contribution from both the second and first order fluctuations in $\bbh{\Psi}$, which now have a non zero expectation value. In this case the bias sums up to 
\begin{equation}
\label{mockscalinguncorrected}
\langle\h{\Sigma}_\bb{p}\rangle=\Sigma_\bb{p}(1-(1+N_b)/N_r+B(1+N_p))
\end{equation}
\item If we repeat the same exercise as above, but we correct for the bias in the precision matrix estimator, we are left with

\end{itemize} 
%
In the three previous cases we only kept the terms of order $O(1/N_r)$. The last two examples have to be considered as academic exercises, as in a real experiment the data has to be used in the end to compute the posterior location parameter and the parameter covariance. With the $\bbh{d}$ Gaussianity assumption, the coefficient $B$ can be calculated exactly as \citep{DodelsonSchneider13,Taylor12}:
\begin{equation}
B = \frac{N_r-N_b-2}{(N_r-N_b-1)(N_r-N_b-4)}
\end{equation}
%
In the limit $N_r\gg N_b$ this leads to a degradation of order $D/N_r$ for the variance of each estimated parameter $\sigma^2_p=\Sigma_{pp}$. We test the scaling relation
\begin{equation}
\label{ourscaling}
\langle\h{\sigma}_p^2\rangle = \sigma^2_p(N_s)\left(1+\frac{D}{N_r}\right)
\end{equation}
%
on our simulations. We call $D$ the \textit{effective dimensionality} of the feature space which, under Gaussianity assumptions, is equal to $D=N_b-N_p$ in the case where the true covariance $\bb{C}$ is known, $D=N_p-N_b$ when the true precision matrix is replaced by its biased estimator (computed from the same simulation set) and $D=1+N_p$ when the $\bbh{\Psi}$ is corrected.

@@

\begin{figure*}
\includegraphics[scale=0.3]{Figures/effective_nb.eps}
\includegraphics[scale=0.3]{Figures/effective_nb_fake_uncorrected.eps}
\includegraphics[scale=0.3]{Figures/effective_nb_fake_corrected.eps}
\caption{Effective dimesionality of the feature space, $D$, as a function of the number of bands $N_b$, for the features outlined in Table \ref{featuretable}. The dimensionality $D$ is plotted in units of its predicted value from equations (\ref{dodelsonscaling}) in the left panel, equation (\ref{mockscalinguncorrected}) in the center panel and equation (\ref{mockscalingcorrected}) in the right panel.}
\label{effectivenb}
\end{figure*}

@@

\bibliography{ref}
\label{lastpage}
\end{document}

>>>>

\begin{abstract}
Constraining cosmology using Weak Gravitational Lensing consists in comparing measured features with simulated ones. An accurate estimate of the feature covariance matrix is essential to obtain accurate parameter confidence intervals. When this covariance matrix $\bb{C}$ is measured from simulations, an important question to ask is how big the simulation set should be for an accurate estimation of $\bb{C}$. We construct different mock ensembles with $N_r$ realizations of the same shear field, based on $N_s<N_r$ independent $N$--body simulations. Using the shear--shear power spectrum as a summary statistic, we find that a single $N$--body simulation can be recycled a large number of times to produce an ensemble of $N_r=10^5$ shear realizations that are mutually independent. We also find that, for a range of summary statistics that include shear--shear power spectra measured at different scales and convergence peak counts, parameter constraints degrade faster than the $O(N_b)$ prediction by \citep{DodelsonSchneider13}, where $N_b$ is the length of the feature vector.       
\end{abstract}

@@

\subsection{Shear field simulations}
\label{shearsim}
%
In this paragraph we describe how we constructed our shear field ensembles. Background galaxies at redshift $z_s$ are lensed by large scale structure between $z=0$ and $z_s$. The shape distortions due to the cosmic shear $\pmb{\gamma}$ can be computed in terms of the dark matter gravitational potential $\Phi(\bb{x},z)$. Because the evolution of $\Phi$ in redshift is non--linear, its evolution needs to be computed with numerical simulations. We make use of the public code \texttt{Gadget2} \citep{Gadget2}, with which we run a sequence of dark--matter--only $N$--body simulations that track the evolution of the density fluctuations in the large scale structure of the universe. We assume a standard $\Lambda$CDM framework with $(\Omega_m,\Omega_\Lambda,h,w,\sigma_8,n_s)=(0.26,0.74,0.72,-1,0.8,0.96)$ and we fix the comoving size of the simulation box to $240\mathrm{Mpc}/h$. We fill the box with $512^3$ particles, which correspond to a mass resolution of $10^{10}M_\odot$. We assume a uniform galaxy distribution at a constant redshift $z_s=2$ (at which the simulation box has an angular size of $\theta_{box}=3.5\mathrm{deg}$) and we disctretize the mass distribution between $z_s$ and the observer at $z=0$ with a sequence of 46 two dimensional lenses of thickness $80\mathrm{Mpc}/h$. The surface density on each lens is computed by projecting the three dimensional density measured from \texttt{Gadget2} snapshots. We then apply the multi--lens--plane algorithm (see \citep{RayTracingHartlap,RayTracingJain} for example) to trace the deflections of $2048^2$ light rays arranged on a square grid of total size $\theta_{box}$, from $z=0$ to $z_s$. The light deflection calculations have been performed with our implementation of multi--lens--plane algorithm, which is part of the \texttt{LensTools} computing package \citep{LensTools} (released under the \texttt{MIT} license). Different realizations $r$ of the same shear field $\pmb{\gamma}_r(\pmb{\theta})$ can be generated with different choices of the lenses that lay between the observer and $z_s$. The randomization procedure we adopt is the following:

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{figure*}
\includegraphics[scale=0.3]{Figures/effective_nb.eps}
\includegraphics[scale=0.3]{Figures/effective_nb_fake_uncorrected.eps}
\includegraphics[scale=0.3]{Figures/effective_nb_fake_corrected.eps}
\caption{Effective dimesionality of the feature space, $D$, as a function of the number of bands $N_b$, for the features outlined in Table \ref{featuretable}. The solid black lines show the theory predictions from equations (\ref{dodelsonscaling}) in the left panel, equation (\ref{mockscalinguncorrected}) in the right panel and equation (\ref{mockscalingcorrected}) in the bottom panel.}
\label{effectivenb}
\end{figure*}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{equation}
\label{covest}
\bbh{C} = \frac{1}{N_r-1}\sum_{r=1}^{N_r} (\bbh{d}_r - \bar{\bb{d}}) (\bbh{d}_r - \bar{\bb{d}})^T
\end{equation}
%
Assuming a normal feature likelihood, together with a flat prior on the parameter space, the parameter posterior distribution $\mathcal{L}(\bb{p}\vert\bbh{d})$ can be obtained using Bayes theorem
\begin{equation}
\label{posteriorbayes}
\mathcal{L}(\bb{p}\vert\bbh{d}) = \mathcal{N}_\mathcal{L}\exp{\left[-\frac{1}{2}(\bbh{d}-\bb{d}(\bb{p}))^T\bb{C}^{-1}(\bbh{d}-\bb{d}(\bb{p}))\right]}
\end{equation}
%
For the sake of simplicity, we approximate the posterior as a Gaussian around its maximum. This is easily done approximating the simulated feature at first order around a point $\bb{p}_0$ (that ideally is the maximum of (\ref{posteriorbayes}))
\begin{equation}
\bb{d}(\bb{p}) \approx \bb{d}_0 + \bb{d}_0^\prime(\bb{p}-\bb{p}_0) 
\end{equation}
%
To measure the feature derivatives $\bb{d}'_0$ with respect to cosmology, we make use of the public code code NICAEA \citep{Nicaea} for the power spectrum, and we use an independent simulation set (that contains mocks at different combinations of $(\Omega_m,w,\sigma_8)$) for the peak counts.
We can build the estimator for the posterior location parameter $\bbh{p}$, given the observation $\bbh{d}$, as follows:
%
\begin{equation}
\label{estimatormean}
\bbh{p} = \bb{p}_0 + \bbh{T}(\bbh{d}-\bb{d}_0)
\end{equation}

\begin{equation}
\label{dodelsonscaling}
\langle\h{\Sigma}_\bb{p}\rangle=\Sigma_\bb{p}(1+B(N_b-N_p))
\end{equation}
%
This is the result obtained by \citep{DodelsonSchneider13}.

\item If we are not interested in measuring the location parameter $\bbh{p}$, but just the width of the parameter posterior, we are tempted to estimate errorbars purely from simulations, replacing the true covariance $\bb{C}$ with its estimator $\bbh{C}$, measured from the same simulation set we use to compute $\bbh{\Psi}$. If we do that, and we do not correct for the bias (\ref{psibias}), the parameter variance will have a contribution from both the second and first order fluctuations in $\bbh{\Psi}$, which now have a non zero expectation value. In this case the bias sums up to 
\begin{equation}
\label{mockscalinguncorrected}
\langle\h{\Sigma}_\bb{p}\rangle=\Sigma_\bb{p}(1-(1+N_b)/N_r+B(1+N_p))
\end{equation}

\item If we repeat the same exercise as above, but we correct for the bias in the precision matrix estimator, we are left with

\end{itemize} 
%
The complete calculations that justify (\ref{dodelsonscaling}),(\ref{mockscalinguncorrected}) and (\ref{mockscalingcorrected}) are shown in the Appendix. The last two examples have to be considered as academic exercises, as in a real experiment the data $\bbh{d}$ and its covariance matrix regulate the scatter of the $\bbh{p}$ estimator. With the $\bbh{d}$ Gaussianity assumption, the coefficient $B$ can be calculated exactly as \citep{DodelsonSchneider13,Taylor12}, and tends to $1/N_r$ in the limit of $N_r\gg N_b$. This leads to a degradation of order $D/N_r$ for the variance of each estimated parameter $\sigma^2_p=\Sigma_{pp}$. We test the scaling relation
\begin{equation}
\label{ourscaling}
\langle\h{\sigma}_p^2\rangle = \sigma^2_p(N_s)\left(1+\frac{D}{N_r}\right)
\end{equation}
%
on our simulations. We call $D$ the \textit{effective dimensionality} of the feature space which, under Gaussianity assumptions, is equal to $D=N_b-N_p$ in the case where the true covariance $\bb{C}$ is known, $D=N_p-N_b$ when the true data covariance is replaced by its estimator (computed from the same simulation set) but the biased $\bbh{\Psi}$ estimator is used, and $D=1+N_p$ when the $\bbh{\Psi}$ bias is corrected.

@@

\bibliography{ref}

%%%%%%%%%%%%%%%%%%%%%%%%%% APPENDIX %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section*{Appendix: scaling of the effective dimensionality}
\label{appendix}

The goal of this appendix is giving a derivation of (\ref{dodelsonscaling}),(\ref{mockscalinguncorrected}) and (\ref{mockscalingcorrected}) following the guidelines from \citep{DodelsonSchneider13,Taylor12}. First consider the case in which we use the same simulation set to estimate $\bb{C},\bbh{\Psi}$: in this case equation (\ref{estimatorcovariance}) reduces to the inverse Fisher estimator $\h{\Sigma}_\bb{p}=\bbh{F}^{-1}$. If we write $\bbh{F}=\bb{F}+\delta\bbh{F}=\bb{F}+\bb{d}_0'^T\delta\bbh{\Psi}\bb{d}_0'$, and expand at second order in the fluctuations we get

\begin{equation}
\h{\Sigma}_\bb{p} = \bb{F}^{-1} + \bb{F}^{-1}\left(- \delta{\bbh{F}}+ \delta{\bbh{F}}\bb{F}^{-1}\delta{\bbh{F}}\right)\bb{F}^{-1}  
\end{equation}  
%
If we use the precision estimator $\bbh{\Psi}=\bb{\Psi}+\delta\bbh{\Psi}$ and do not correct for the bias (\ref{psibias}), we have to use the facts that
\begin{equation}
\begin{matrix}
& \langle\delta\bbh{\Psi}\rangle = \frac{1+N_b}{N_r-N_b-2}\bb{\Psi} \\ \\
& \langle\delta\h{\Psi}_{ij}\delta\h{\Psi}_{kl}\rangle = \frac{A([N_r-1)^2+(1+N_b)^2]\Psi_{ij}\Psi_{kl} + B(N_r-1)^2(\Psi_{ik}\Psi_{kl}+\Psi_{il}\Psi_{jk})}{(N_r-N_b-2)^2}
\end{matrix}
\end{equation}
%
Keeping only the leading order $O(1/N_r)$ in the expansion we get the results
\begin{equation}
\label{explanationuncorrected}
\begin{matrix}
& \langle\delta\bbh{F}\rangle = \left(\frac{1+N_b}{N_r}\right)\bb{F} \\ \\
& \langle\delta{\bbh{F}}\bb{F}^{-1}\delta{\bbh{F}}\rangle = B(1+N_p)\bb{F}
\end{matrix}
\end{equation}
%
which leads immediately to (\ref{mockscalinguncorrected}). Now suppose we correct for the bias in the precision estimator, and use the unbiased $\bbh{\Psi}$ for which $\delta\bbh{\Psi}=0$. This leads to equation (\ref{psifluctuations}) for the quadratic fluctuations in the precision matrix, and to the same (\ref{explanationuncorrected}), except for $\langle\delta\bbh{F}\rangle=0$. This leads to (\ref{mockscalingcorrected}). If now we keep $\bb{C}$ fixed and we estimate only $\bbh{\Psi}$ from the simulations, we get
\begin{equation}
\label{tempcovarianceestimate}
\h{\Sigma}_\bb{p} = (\bbh{F}+\delta\bbh{F})^{-1}\bb{d}_0'^T(\bbh{\Psi}+\delta\bbh{\Psi})\bb{C}(\bbh{\Psi}+\delta\bbh{\Psi})\bb{d}_0'(\bbh{F}+\delta\bbh{F})^{-1}
\end{equation}
%
If we isolate the second order terms in (\ref{tempcovarianceestimate}), the result is 
\begin{equation}
\h{\Sigma}_\bb{p}\vert_2 = \bb{F}^{-1}(\bb{d}_0'^T\delta\bbh{\Psi}\bb{C}\delta\bbh{\Psi}\bb{d}_0'-\delta{\bbh{F}}\bb{F}^{-1}\delta{\bbh{F}})\bb{F}^{-1}
\end{equation} 
%
The expectation values of the quadratic fluctuations are easily computed using (\ref{psifluctuations}):
\begin{equation}
\begin{matrix}
& \langle\bb{d}_0'^T\delta\bbh{\Psi}\bb{C}\delta\bbh{\Psi}\bb{d}_0'\rangle = B(1+N_b)\bb{F} \\ \\
& \langle\delta{\bbh{F}}\bb{F}^{-1}\delta{\bbh{F}}\rangle = B(1+N_p)\bb{F}
\end{matrix}
\end{equation}
%
From which (\ref{dodelsonscaling}) immediately follows. 
\label{lastpage}
\end{document}
